Exploring mathematical objects from custom-tailored mathematical universes

### 2020-03-28 paper-filmat.tex
Details on the injection ℝ → ℕ, prompted by Franka Waaldijk
MO thread: https://mathoverflow.net/questions/355936/contrasting-theorems-in-classical-logic-and-constructivism/

The realizability topos given by infinite-time Turing machines provides an intriguing environment challenging many
mathematical intuitions shaped by classical logic. For instance, while from the
point of view of this topos the reals are still uncountable in the sense that
there is no surjection~$\NN \to \RR$, there is an injection~$\RR \to
\NN$~\cite{bauer:injection}.

>>>>

The realizability topos given by infinite-time Turing machines provides an intriguing environment challenging many
mathematical intuitions shaped by classical logic. For instance, while from the
point of view of this topos the reals are still uncountable in the sense that
there is no surjection~$\NN \to \RR$, there is an injection~$\RR \to
\NN$~\cite{bauer:injection}.\footnote{What the realizability topos
given by infinite-time Turing machines believes to be a real number is, from
the external point of view, an infinite-time Turing machine~$X$ which outputs,
when called with a natural number~$n$ as input, a rational
approximation~$X(n)$. As with the original \effective topos, these
approximations have to be consistent in the sense that~$|X(n) - X(m)| \leq
2^{-n} + 2^{-m}$, and two such machines~$X, X'$ represent the same real
iff~$|X(n) - X'(m)| \leq 2^{-n} + 2^{-m}$ for all natural numbers~$n,m$.

A map~$\RR \to \NN$ in this topos is hence given by an infinite-time
Turing machine~$M$ which, when given the description of such an infinite-time
Turing machine~$X$ as input, outputs a certain natural number~$M(X)$. If~$X$
and~$X'$ represent the same real, then~$M(X)$ has to coincide with~$M(X')$.
This map~$\RR \to \NN$ is injective iff conversely,~$M(X) = M(X')$ implies
that~$X$ and~$X'$ represent the same real.

We can program such a machine~$M$ as follows: Read the description of an
infinite-time Turing machine~$X$ representing a real number as input. Then
simulate, in a dovetailing fashion, all infinite-time Turing machines and
compare their outputs with the outputs of~$X$. As soon as a machine~$X'$ is
found which happens to terminate on all inputs in such a way that~$|X(n) -
X'(m)| \leq 2^{-n} + 2^{-m}$ for all natural numbers~$n,m$, output the number
of this machine (in the chosen enumeration of all infinite-time Turing
machines) and halt.

The number~$M(X)$ computed by~$M$ depends on the input/output behaviour of~$X$,
the chosen order of infinite-time Turing machines, and on details of the
interleaving simulation and comparison procedure -- but it does not depend on
the implementation of~$X$ or on its specific choice of rational
approximations~$X(n)$. The search terminates since there is at least one
infinite-time Turing machine which represents the same real number as~$X$ does,
namely~$X$ itself.}
